---
title: "epi curious"
author: "Ania Kawiecki"
date: "9/29/2020"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```

```{r libraries, message= FALSE}

library(tidyverse)
library(rethinking)
library(dagitty)
library(knitr)
library("ggdag")
library("ggrepel")

library(INLA)
library(knitr)
library(stringr)
library("patchwork")
library("ghibli")
library("fpp2")
```

VECTOR BORNE DISEASE

Table of 20 most important VB diseases in the world: where when and what (pathogen, vector, geo dist) 
Vectorial capacity, vector competence, host competence.
How do you assess a reservoir? 
1. Pathogen is present in the host in nature.
 If you experimentally infect that animal it gets infected. ?
Main strategies of vector intervention: treatments, water removal, larvicidal 
Ross-McDonalds model: boils parameters down to the components of transmission. They also set out an intervention strategy by intervention on the parameter. Ex development time in the host and vector. 
Compare contrast vectors mosquito vs ticks. Ex: ticks attach and inject salivary proteins. Mosquitoes feed repeatedly and quickly 
Read some medical entomology
Brush up on vector indexes, especially for adult mosquitoes. Ex aedes adult house index. 

Vectorial capacity, vector competence, host competence.

